---
title: Surplus production
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

Production functions:

```{r}
fletcher <- function(B, Binf, m) {
  (4 * m) / Binf * (1 - B / Binf) * B
}
B <- 1:100
plot(B, sapply(1:100, function(x) fletcher(x, 100, 60)), ylab = "Production rate")

schaefer <- function(B, Binf, r) {
  r * (1 - B / Binf) * B
}
B <- seq(0, 100, length.out = 1000)

# find matching r:
get_r <- function(m, Binf) (4 * m) / Binf
get_r(60, 100)

Pdot <- sapply(B, function(x) schaefer(x, 100, 2.4))
plot(B, Pdot, type = "l", ylim = c(-10, max(Pdot)), ylab = "Rate (1/t)", xlab = "B (biomass)")
abline(h = 0)
abline(v = 100 / 2)

# Fm <- (2 * 60) / 100 # after eq 2.8c
# Fm

F <- 0.6
Ydot <- F * B
lines(B, Ydot, col = "blue")

Bstar <- B[min(which(Pdot < Ydot))]
Pdot_star <- Pdot[min(which(Pdot < Ydot))]
Fstar <- Pdot_star / Bstar # 2.4b
Fstar

Ydot_star <- Fstar * B
Bdot <- Pdot - Ydot_star
lines(B, Bdot, col = "red")
abline(v = Bstar)

legend("topleft",
  col = c("black", "blue", "red"),
  legend = c("Pdot (production rate)", "Ydot (yield rate)", "Bdot (biomass rate)"),
  lty = c(1, 1, 1)
)
```

Simulate from a Schaefer surplus production model:

```{r, fig.asp=1.2}
set.seed(1)
sim_schaefer <- function(K, r, q, sigma, N, C) {
  B <- numeric(N)
  U <- numeric(N)
  B[1] <- K
  U[1] <- q * B[1]
  for (i in 2:N) {
    B[i] <- B[i - 1] + r * B[i - 1] * (1 - (B[i - 1] / K)) - C[i - 1]
    U[i] <- q * B[i]
  }
  index <- exp(log(U) + rnorm(N, -(sigma^2) / 2, sigma))
  list(index = index, catch = C, B = B)
}

catches <- c(seq(0, 96, length.out = 50), seq(96, 0, length.out = 50))
sim_dat <- sim_schaefer(K = 1000, r = 0.3, q = 0.1, sigma = 0.2, N = length(catches), C = catches)
par(mfrow = c(3, 1))
plot(sim_dat$index, type = "o")
plot(sim_dat$catch, type = "o")
plot(sim_dat$B, type = "o")
```

A function to return the negative log likelihood or predictions:

```{r}
sp_nll <- function(par, dat, return_predictions = FALSE) {
  N <- dat$N
  catch <- dat$catch
  Uobs <- dat$Uobs
  
  K <- exp(par["log_K"])
  r <- exp(par["log_r"])
  q <- exp(par["log_q"])
  sigma <- exp(par["log_sigma"])
  
  B <- numeric(N)
  U <- numeric(N)
  B[1] <- dat$initial_depletion * K
  for (i in 2:N) {
    B[i] <- B[i - 1] + r * B[i - 1] * (1 - (B[i - 1] / K)) - catch[i - 1]
  }
  U <- q * B # vectors
  U[U < 0.0001] <- 0.0001
  
  nll <- 0
  # # prior on r:
  # nll <- -dnorm(log(r), log(0.3), sd = 0.3, log = TRUE)
  
  # # prior on K:
  # nll <- nll - dnorm(log(K), log(1000), sd = 0.1, log = TRUE)
  
  # observation likelihood on index:
  nll <- nll - sum(dnorm(log(U), log(Uobs) - (sigma^2) / 2, sd = sigma, log = TRUE))
  
  if (!return_predictions) {
    return(nll)
  } else {
    return(list(U = U, B = B))
  }
}
```

```{r}
# create input data:
dat <- list()
dat$catch <- sim_dat$catch
dat$Uobs <- sim_dat$index
dat$N <- length(sim_dat$catch)
dat$initial_depletion <- 1

# initial values:
pars <- c(log_K = log(1000), log_r = log(0.3), log_q = log(0.1), log_sigma = log(0.2))

# test it:
sp_nll(pars, dat = dat)

# fit it:
fit <- nlminb(pars, sp_nll, control = list(eval.max = 1000, iter.max = 1000), dat = dat)

par_hat <- unname(exp(fit$par))
names(par_hat) <- gsub("log_", "", names(pars))
round(par_hat, 2)

est <- sp_nll(fit$par, dat = dat, return_predictions = TRUE)
par(mfrow = c(2, 1))
plot(est$U, type = "l", ylab = "Index", xlab = "Year")
points(sim_dat$index)
plot(est$B, type = "l", ylab = "Biomass", xlab = "Year")
lines(sim_dat$B, lty = 2)
```

That was a "two-way trip" with moderately strong fishing, which is quite informative. Try others.

First, make function to do the fitting and plotting:

```{r}
fit_sp <- function(catches, K = 1000, r = 0.3, q = 0.1, sigma = 0.2, title = "") {
  sim_dat <- sim_schaefer(K = K, r = r, q = q, sigma = sigma, N = length(catches), C = catches)
  cat("True pars: K = ", K, ", r = ", r, ", q =", q, ", sigma = ", sigma, "\n", sep = "")
  par(mfrow = c(3, 1))
  plot(sim_dat$catch, type = "o", ylab = "Catch", xlab = "Year", main = title)
  
  dat <- list()
  dat$catch <- sim_dat$catch
  dat$Uobs <- sim_dat$index
  dat$N <- length(sim_dat$catch)
  dat$initial_depletion <- 1
  
  pars <- c(log_K = log(K), log_r = log(r), log_q = log(q), log_sigma = log(sigma))
  fit <- nlminb(pars, sp_nll, control = list(eval.max = 2000, iter.max = 2000), dat = dat)
  
  par_hat <- unname(exp(fit$par))
  names(par_hat) <- gsub("log_", "", names(pars))
  cat("Estimated pars:\n")
  print(round(par_hat, 2))
  
  est <- sp_nll(fit$par, dat = dat, return_predictions = TRUE)
  plot(est$U, type = "l", ylab = "Index", xlab = "Year", ylim = range(est$U, sim_dat$index))
  points(sim_dat$index)
  plot(est$B, type = "l", ylab = "Biomass", xlab = "Year", ylim = range(c(est$B, sim_dat$B)))
  legend("topleft", legend = c("True", "Estimated"), lty = c(2, 1))
  lines(sim_dat$B, lty = 2)
}
```

Two-way trip:

```{r, fig.asp=1.3}
catches <- c(seq(0, 96, length.out = 50), seq(96, 0, length.out = 50))
fit_sp(catches, title = "One-way trip")
fit_sp(catches, title = "One-way trip")
```

Lightly fished:

```{r, fig.asp=1.3}
set.seed(123)
catches <- c(seq(0, 12, length.out = 50), seq(12, 0, length.out = 50))
fit_sp(catches, title = "Lightly fished")
fit_sp(catches, title = "Lightly fished")
fit_sp(catches, title = "Lightly fished")
```

One-way trip:

```{r, fig.asp=1.3}
catches <- c(seq(0, 95, length.out = 100))
fit_sp(catches, title = "One-way trip")
fit_sp(catches, title = "One-way trip")
```

Initially fished hard:

```{r, fig.asp=1.3}
catches <- c(seq(97, 0, length.out = 100))
fit_sp(catches, title = "Initially fished hard")
fit_sp(catches, title = "Initially fished hard")
```

Roller coaster:

```{r, fig.asp=1.3}
catches <- sin(2 * pi * seq(-1, 1, length.out = 100)) * 20 + 70
fit_sp(catches, title = "Roller coaster")
```

Roller coaster with noisy survey:

```{r, fig.asp=1.3}
catches <- sin(2 * pi * seq(-1, 1, length.out = 100)) * 20 + 70
fit_sp(catches, title = "Roller coaster", sigma = 0.8)
fit_sp(catches, title = "Roller coaster", sigma = 0.8)
fit_sp(catches, title = "Roller coaster", sigma = 0.8)
```

# Do it in Stan

```{txt, code=readLines('schaefer_log.stan')}
```

```{r, stan-fit, warning=FALSE, message=FALSE}
library(cmdstanr)

catches <- c(seq(0, 96, length.out = 50), seq(96, 0, length.out = 50))
K <- 1000
r <- 0.3
q <- 0.2
sigma <- 0.2

set.seed(1)
sim_dat <- sim_schaefer(K = K, r = r, q = q, sigma = sigma, N = length(catches), C = catches)
par(mfrow = c(1, 1))
plot(sim_dat$catch, type = "o", ylab = "Catch", xlab = "Year")

dat <- list(
  C = sim_dat$catch,
  log_Uobs = log(sim_dat$index),
  init_depletion = 1,
  log_K_mean = log(K),
  log_q_mean = log(q),
  log_r_mean = log(r),
  log_K_sd = 2,
  log_q_sd = 1,
  log_r_sd = 1,
  N = length(sim_dat$catch)
)

init <- lapply(seq_len(4), function(i) {
  list(
    log_K = rnorm(1, log(K), 0.5), sigma = rlnorm(1, log(sigma), 0.1),
    log_q = rnorm(1, log(q), 0.5), log_r = rnorm(1, log(r), 0.5)
  )
})

mod <- cmdstan_model("schaefer_log.stan")
fit <- mod$sample(
  data = dat, init = init, parallel_chains = 4L,
  iter_sampling = 2000, iter_warmup = 1000, chains = 4, seed = 123,
  adapt_delta = 0.99, max_treedepth = 15L
)
fit$cmdstan_diagnose()
stanfit <- rstan::read_stan_csv(fit$output_files())
```

```{r}
print(stanfit, pars = c("log_K", "log_q", "log_r", "sigma"))
pairs(stanfit, pars = c("log_K", "log_q", "log_r", "sigma"))
p <- rstan::extract(stanfit)
par(mfrow = c(2, 2))
hist(exp(p$log_K));abline(v = K, col = "red", lwd = 1.5)
hist(exp(p$log_q));abline(v = q, col = "red", lwd = 1.5)
hist(exp(p$log_r));abline(v = r, col = "red", lwd = 1.5)
hist(p$sigma);abline(v = sigma, col = "red", lwd = 1.5)
```
