---
title: Pella Tomlinson *m* exploraiton
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
```

*I'll use Q&D and H&W as shorthand for Quinn & Deriso 1999 and Hilborn & Walters 1992*

Copy over the Fletcher from production-models.Rmd  
Fletcher's formulation is used in subsequent development because of the inherent utility of its parameters (Quinn & Deriso).  
```{r}
fletcher <- function(B, Binf, m) {
  (4 * m) / Binf * (1 - B / Binf) * B
}
B <- 1:100
plot(B, type = "l", sapply(1:100, function(x) fletcher(x, 100, 60)), ylab = "Production rate")
```
  
  *How does this  m relate to Pella-Tomlinson in Hilborn and Walters (1992) Eq. 8.4.4?*  
**m* in Q&D is sub notation meaning maximum, but also units? See Eq 2.5 and following text.  
*If $m = 60$ in the equation above, and Q&D call this the maximum, why is the peak at 50?

Now the Pella-Tomlinson independent of catch. With *m* = 2 so it should look like above.  
*could be cool to write this to take multiple m*
```{r}
PelTom <- function(B, r, k, m) {
  (r*B) - ((r/k)*B^m)
}
Pdot <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2)) #same B as above
plot(B, Pdot, type = "l", ylim = c(-10, max(Pdot)), ylab = "Rate (1/t)", xlab = "B (biomass)")
```
  
How does changing *m* affect this shape of productivity? 
```{r}
Pdotm1 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 1)) #same B as above
Pdotm2 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2)) #same B as above
Pdotm3 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 3)) #same B as above

plot(Pdotm1, type = "l", ylab = "Rate (1/t)", xlab = "B (biomass)", col = "blue")
lines(B, Pdotm2, col = "black")
lines(B, Pdotm3, col = "red")
```
  
Try again with more realistic m??
```{r}
Pdotm1 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 1.5)) #same B as above
Pdotm2 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2)) #same B as above
Pdotm3 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2.5)) #same B as above

plot(Pdotm1, type = "l", ylab = "Rate (1/t)", xlab = "B (biomass)", col = "blue")
lines(B, Pdotm2, col = "black")
lines(B, Pdotm3, col = "red")
```
  
Almost looks the same but look at the red line!  
Try again with different values. 
```{r}
B <- 1:180 #adjust range of biomass so we can see more

Pdotm1 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 1.9)) #same B as above
Pdotm2 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2)) #same B as above
Pdotm3 <- sapply(B, function(x) PelTom(x, r = 2.4, k = 100, m = 2.1)) #same B as above

plot(Pdotm1, type = "l", ylab = "Rate (1/t)", xlab = "B (biomass)", col = "blue")
lines(B, Pdotm2, col = "black")
lines(B, Pdotm3, col = "red")
abline(h = 0)
```
