---
title: Recruitment exploraiton
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
```
## Beverton-Holt  
Expnded (eq 3.6) where `a` and `b` are density dependent, and independent additive effects on mortality, `Z`, and `T` is age at recruitment.    
```{r long beverton-holt}
S <- 1:100 #vector of spawners
T <- 4 #age @ recruitment
f <- 10  #avg net fecundity
a1 <- 0.5 #DI effect on mortality
b1 <- 0.1 #DD effect on mortality

R_L_BH <- S/((exp(a1*T)/f) + (b1/a1)*(exp(a1*T) - 1)*S)

plot(R_L_BH~S)
```
  
We can get the more normal parameters, alpha and beta, with some math.  
- What do a1 and b1 *really mean*? How would one choose a reasonable starting value for them for a population? Can you empirically estimate these? Or they more abstract?  
```{r}
alpha <- f*exp(-a1*T)
beta <- f*b1*(1-exp(-a1*T))/a1
```
Then plug these values into the Beverton-Holt and get the same figure as earlier.   
```{r beverton holt}
R_BH <- (alpha*S)/(1+beta*S)

plot(R_BH~S)
```
  
### Ricker
Now we do the same as above by formulating the Ricker (eq. 3.8) with additive DI and DD effects on mortality. **I'm going to try keeping the a1 and b1 as above, but I'm not sure this is proper**  
- It's probably not proper because  they use *a2* and *b2* (eq 3.7); these still mean the same thing but they need to be different numbers?   
```{r long Ricker}
#S <- seq(0, 10, by = 0.1) # helper to toggle S around
b1 <- 0.01 #overwrite b to make a nice fig 
R_L_R <- f*exp(-a1*T)*S*exp(-b1*T*S)

plot(R_L_R ~ S) #my code & math is right I just picked weird values for a and B I think... 
```

If everything is right I *should be* able to write the Ricker with the alpha and betas I calculated earlier?  
- No. `a2` and `b2` behave differently in the equation. See top line of eq 3.8 and compare with eq 3.6 while looking at in-text math for `a` and `b`. 
- Is there any utility in isolating alpha and beta to find values for `a2` and `b2` like we did above for the Beverton-Holt?   
```{r regular ricker}
R_R <- alpha*S*exp(-beta*S)

plot(R_R ~ S, main = "bad Ricker!") #nope! - this should match the previous fig. 
```
  
We'll go ahead and pick nice parameters to use later, and plot a second Ricker. 
```{r}
alpha <- 0.5
  
beta <- 0.05
  
R_R2 <- alpha*S*exp(-beta*S)

plot(R_R2~S, main = "better Ricker")
```
  
### Ludwig-Walters model  
where the DD mortality term is a power function of the Spawning stock  
If the 3rd parameter, `gamma`,  = 1 it's a Ricker. They suggest setting `gamma` = 2  
- Is this parameter fit or fixed? Seems like it could be hard to estimate.  
```{r}
gamma <- 2
R_LW <- alpha*S*exp(-beta*S^gamma)

plot(R_LW~S)
```
  
### Environmental variation  
Other *normally distributed* environmental effects also influence recruitment, and can be incorporated into the exponent of a Ricker (eq 3.10), or in a linear Ricker in the form of multiple regression. The point is that we add stochastic variation to the Ricker via lognormally distributed error, first by generating random, normal noise, then exponentiating it in the Ricker.    
```{r, Ricker with error}
proc_error <- rnorm(length(S), sd = 0.5)

R_stoch <- alpha*S*exp(-beta*S)*exp(proc_error)

plot(R_stoch ~ S)
lines(R_R2 ~ S)
```
### Cushing  
Two parameter model not used much in practice. *Why show the other parameterizaiton here?*    
```{r, cushing}
gamma <- 0.5
R_cush <- alpha*S^gamma

plot(R_cush ~S, main = paste("cushing where gamma =", gamma))

plot((alpha*S^1)~S, main = "cushing with different gammas")
lines((alpha*S^0.8)~S)
lines((alpha*S^1.2)~S)
```

### Deriso-Schnute  
Three parameter model based on 2 species considering predator-prey. Added comments for Schnute's definitions of parms.    
When `gamma <- -1` you have a Beverton-Holt, a Ricker as `gamma` approaches 0. As `gamma` approaches `Inf` recruitment becomes proportional to the spawning stock.  
*Skipping derivation steps*
```{r, D-S}
#toggle as you please
alpha <- 0.5 #the productivity parameter
beta <- 0.05 #the optimality parameter
gamma <- 0.5 #recruitment limitation/skewness parameter

R_DS <- alpha*S*(1-(beta*gamma*S))^(1/gamma)


plot(R_DS~S, main = paste("Deriso-Schnute(alpha=", alpha, "beta=", beta, "gamma=", gamma, ")"))
```
  
#### Figure 3.5  
- Hmmm, need to pick better numbers?
```{r, 3.5}
plot((alpha*S*(1-(beta*1*S))^(1/1))~S, main = "cushing with different gammas")
lines((alpha*S*(1-(beta*-1*S))^(1/-1))~S)
lines((alpha*S*(1-(beta*0*S))^(1/0))~S)
```
  
### Shepard
When `gamma <- 1` you have a Beverton-Holt, when `gamma > 1` a dome shaped SR curve, and when `gamma < 1` a curve that increases indefinitely like the Cushing.   
```{r, Shepard}

```
